Social inequality in mortality among adults and elderly in northern Sweden 1851--2013\footnote{Paper presented at the session 'Health and Mortality in Arctic Populations'. The Ninth International Congress on Arctic Social Sciences, Umeå, June 8--12, 2017}

### 2017-06-01 icass.Rnw
add tables, natbib with apalike

@@

\bibliographystyle{plainnat}

@@

%\usepackage[tablesonly]{endfloat}
%\usepackage[figuresonly]{endfloat}
%\usepackage[tablesonly]{endfloat}

%\renewcommand{\efloatseparator}{\vfill}

% patch \efloat@iwrite to use \protected@write
%\makeatletter
%\renewcommand\efloat@iwrite[1]{%
%   \immediate\expandafter\protected@write\csname efloat@post#1\endcsname{}}
%\makeatother

@@

@@

<<setup, include = FALSE>>=
library(knitr)
opts_chunk$set(engine='R',dev='pdf',fig.path='figs/slide-',tidy=FALSE,fig.height=5.5,fig.width=8,strip.white=TRUE,cache=TRUE,echo=FALSE,message=FALSE,comment=NA,show.signif.stars = FALSE)##$
library(knitr)
library(eha)
library(xtable)
library(skum)
@

@@

<<cumsw, fig.cap = cap0, fig.scap = scap0, fig.height = 10>>=
source("R/plotED.glm.R")
oldpar <- par(mfrow = c(2, 1))
woutp <- plotED.glm(resw, col = scol, logScale = FALSE, ylab = "TMR(65|40)",
                main = "Women 40-64", exclude = 5:6)
moutp <- plotED.glm(resm, col = scol, logScale = FALSE, ylab = "TMR(65|40)",
                main = "Men 40-64")##, exclude = 5:6)
par(oldpar)
@

@@

<<cumswstand, fig.cap = cap12, fig.scap = scap12, fig.height = 10>>=
oldpar <- par(mfrow = c(2, 1))
woutp <- plotED.glm(resw, col = scol, logScale = FALSE, relative = TRUE,
                    ylab = "TMR(65|40)", main = "Women 40-64", exclude = 5:6)
moutp <- plotED.glm(resm, col = scol, logScale = FALSE, relative = TRUE,
                    ylab = "TMR(65|40)", main = "Men 40-64")##, exclude = 5:6)
par(oldpar)
@

@@

<<cumswold, fig.cap = cap22, fig.scap = scap22, fig.height = 10>>=
##source("R/plotED.R")
resw <- getPHglm2(vb6589[vb6589$sex == "female", ],
            nper, labb)
resm <- getPHglm2(vb6589[vb6589$sex == "male", ],
            nper, labb)

@@

@@

<<41cumsw, fig.cap = cap12, fig.scap = scap12, fig.height = 10>>=
source("R/plotED.glm.R")
oldpar <- par(mfrow = c(2, 1))
woutp <- plotED.glm(resw, col = scol, logScale = FALSE, relative = TRUE, exclude = 3:4,
                    ylab = "TMR(65|40)",  main = "Women 40-64")
moutp <- plotED.glm(resm, col = scol, logScale = FALSE, relative = TRUE, ##exclude = 1:2,
                    ylab = "TMR(65|40)", main = "Men 40-64")
par(oldpar)
@

@@

<<41cumswold, fig.cap = cap22, fig.scap = scap22, fig.height = 10>>=
##source("R/plotED.R")
oldpar <- par(mfrow = c(2, 1))
woutp <- plotED.glm(resw, col = scol, logScale = FALSE, relative = TRUE,
                    ylab = "H(90|65)",  main = "Women 65-89", exclude = 3:4)
moutp <- plotED.glm(resm, col = scol, logScale = FALSE, relative = TRUE,
                    ylab = "H(90|65)", main = "Men 65-89")
par(oldpar)
@

@@

<<12cumsw, fig.cap = cap12, fig.scap = scap12, fig.height = 10>>=
source("R/plotED.glm.R")
oldpar <- par(mfrow = c(2, 1))
woutp <- plotED.glm(resw, col = scol, logScale = FALSE, relative = TRUE, exclude = 3:4,
                    ylab = "TMR(65|40)",  main = "Women 40-64")
moutp <- plotED.glm(resm, col = scol, logScale = FALSE, relative = TRUE, ##exclude = 1:2,
                    ylab = "TMR(65|40)", main = "Men 40-64")
par(oldpar)
@

@@

@@

<<12cumswold, fig.cap = cap22, fig.scap = scap22, fig.height = 10>>=
##source("R/plotED.R")
oldpar <- par(mfrow = c(2, 1))
woutp <- plotED.glm(resw, col = scol, logScale = FALSE, relative = TRUE,
                    ylab = "TMR(90|65)",  main = "Women 65-89", exclude = 3:4)
moutp <- plotED.glm(resm, col = scol, logScale = FALSE, relative = TRUE,
                    ylab = "TMR(90|65)", main = "Men 65-89")
par(oldpar)
@

% \subsubsection{The effect of income and education on cancer death, 1990--2005}
% 
% Thus, we find indications of social differences in the later decades, particularly between the elite and the workers. The alternative analysis for the last period is based on a separate analysis with
% income and education as explanatory variables.
% 
% <<12lissyoung2>>=
% library(eha)
% library(xtable)
% ##load("data/loTab.rda")
% ##loTab$period <- factor(loTab$period)
% liss6589 <- loTab[loTab$age > 64.5, ]
% @
% 
% 
% The statistical significance of the covariates are shown in Table~\ref{tab:12liswom2} for women and Table~\ref{tab:12lismen2} for men.
% <<12liswom2, results = 'asis'>>=
% fitw2 <- glm(event ~ offset(log(exposure)) + as.factor(age) + urban + period + civst + income + educ, family = poisson, data = liss6589[liss6589$sex == "female", ])
% ##cat("Women:\n")
% drw2 <- drop1(fitw2, test = "Chisq")
% xtable(drw2, caption = "Effect of covariates for women 65-89, 1990-2005, cancer death.", label = "tab:12liswom2")
% ##round(summary(fitw1)$coef[-(1:13), ], 4)
% @
% 
% <<12lismen2, results='asis'>>=
% fitm2 <- glm(event ~ offset(log(exposure)) + as.factor(age) + urban + period + civst + income + educ, family = poisson, data = liss6589[liss6589$sex == "male", ])
% ##cat("\nMen:\n")
% drm2 <- drop1(fitm2, test = "Chisq")
% xtable(drm2, caption = "Effect of covariates for men 65-89, 1990-2005, cancer death.", label = "tab:12lismen2")
% ##round(summary(fitm1)$coef[-(1:13), ], 4)
% dcol <- c("blue", "darkgreen", "darkgreen", "red")
% @
% 
% The estimated effect sizes are shown in Figure \ref{fig:12liswomfig2}
% 
% <<12liswomfig2, fig.cap = "Effect sizes, women 65-89, 1990-2005, cancer death.", fig.height = 4>>=
% #source("R/plotWEduc.R")
% plotWEduc(fitw2)
% @
% 
% 
% And for men:
% 
% %% <<12lismen2, results='asis'>>=
% %% fitm2 <- glm(event ~ offset(log(exposure)) + as.factor(age) + urban + period + civst + income + educ, family = ##poisson, data = liss6589[liss6589$sex == "male", ])
% %% ##cat("\nMen:\n")
% %% drm2 <- drop1(fitm2, test = "Chisq")
% %% xtable(drm2, caption = "Effect of covariates for men 65-89, 1990-2005, cancer death.")
% %% round(summary(fitm1)$coef[-(1:13), ], 4)
% %% dcol <- c("blue", "darkgreen", "darkgreen", "red")
% %% @
% 
% <<12lismanfig2, fig.cap = "Effect sizes, men 65-89, 1990-2005, cancer death.", fig.height = 4>>=
% #source("R/plotWEduc.R")
% plotWEduc(fitm2)
% @
% 
% The social inequality in cancer mortality is confirmed by the approach of analysing the effect of education and income. We find the same consistent pattern as we have found for all-cause mortality and cardiovascular mortality for this period. Higher levels of education and more income is associated with lower cancer mortality. We can therefore conclude that both education and income are strongly decisive aspects of social differentiation when it comes to mortality during the last decades.

@@

@@

@@

@@

@@

%The overall pattern is the same for cardiovascular mortality, high income pays off.
<<plotincheart, fig.cap="Cumulative hazards by income and sex, cardiovascular death, 1990--2013.">>=
oldpar <- par(mfrow = c(1, 2))

@@

@@

>>>>

\usepackage{float}

@@

\bibliographystyle{apalike}
%\bibliographystyle{plainnat}

@@

\usepackage[tablesonly]{endfloat}
%\usepackage[figuresonly]{endfloat}
%\usepackage[tablesonly]{endfloat}

\renewcommand{\efloatseparator}{\vfill}

% patch \efloat@iwrite to use \protected@write
\makeatletter
\renewcommand\efloat@iwrite[1]{%
   \immediate\expandafter\protected@write\csname efloat@post#1\endcsname{}}
\makeatother

@@

<<setup, include = FALSE>>=
library(knitr)
opts_chunk$set(engine='R',dev='pdf',fig.path='figs/slide-',tidy=FALSE,fig.height=3.5,fig.width=8,strip.white=TRUE,cache=TRUE,echo=FALSE,message=FALSE,comment=NA,show.signif.stars = FALSE,fig.pos='ht')##$
library(knitr)
library(eha)
library(xtable)
library(skum)
@

@@

@@

<<cumsw, fig.cap = cap0, fig.scap = scap0, fig.height = 8>>=
source("R/plotED.glm.R")
oldpar <- par(mfrow = c(2, 1))
woutp <- plotED.glm(resw, col = scol, logScale = FALSE, ylab = "TMR(65|40)",
                main = "Women 40-64", exclude = 5:6)
moutp <- plotED.glm(resm, col = scol, logScale = FALSE, ylab = "TMR(65|40)",
                main = "Men 40-64")##, exclude = 5:6)
par(oldpar)
@

From Table~\ref{tab:cumswtab} it is obvious that the statistical significance is strong.
The most interesting feature is the three-way interaction with sex, period, and hisclass.

<<cumswtab,results='asis',cap="Table">>=
fitt <- glm(event ~ offset(log(exposure)) + urban + civst + age + sex * hisclass * period, 
            data = vb4064, family = poisson)
dr <- drop1(fitt, test = "Chisq")
##knitr::kable(dr)
xtable(dr,label = "tab:cumswtab", caption = "Anova, all causes of death, 40-65.")
@

@@

<<cumswstand, fig.cap = cap12, fig.scap = scap12, fig.height = 8>>=
oldpar <- par(mfrow = c(2, 1))
woutp <- plotED.glm(resw, col = scol, logScale = FALSE, relative = TRUE,
                    ylab = "TMR(65|40)", main = "Women 40-64", exclude = 5:6)
moutp <- plotED.glm(resm, col = scol, logScale = FALSE, relative = TRUE,
                    ylab = "TMR(65|40)", main = "Men 40-64")##, exclude = 5:6)
par(oldpar)
@

@@

<<cumswold, fig.cap = cap22, fig.scap = scap22, fig.height = 8>>=
##source("R/plotED.R")
resw <- getPHglm2(vb6589[vb6589$sex == "female", ],
            nper, labb)
resm <- getPHglm2(vb6589[vb6589$sex == "male", ],
            nper, labb)

@@

See Table~\ref{tab:cumswoldtab}.

<<cumswoldtab,results='asis'>>=
fitt <- glm(event ~ offset(log(exposure)) + urban + civst + age + sex * hisclass * period, 
            data = vb6589, family = poisson)
dr <- drop1(fitt, test = "Chisq")
##knitr::kable(dr)
xtable(dr,label = "tab:cumswoldtab", caption = "Anova, all causes of death, 65-89.")
@

@@

<<41cumsw, fig.cap = cap12, fig.scap = scap12, fig.height = 8>>=
source("R/plotED.glm.R")
oldpar <- par(mfrow = c(2, 1))
woutp <- plotED.glm(resw, col = scol, logScale = FALSE, relative = TRUE, exclude = 3:4,
                    ylab = "TMR(65|40)",  main = "Women 40-64")
moutp <- plotED.glm(resm, col = scol, logScale = FALSE, relative = TRUE, ##exclude = 1:2,
                    ylab = "TMR(65|40)", main = "Men 40-64")
par(oldpar)
@

%%<<cumsm, fig.cap = "Total haz of deaths in ages 40--64 per individual by HISCLASS and decade, men.">>=
%%@
<<41cumswtab,results='asis'>>=
fitt <- glm(event ~ offset(log(exposure)) + urban + civst + age + sex * hisclass + 
                sex * period + hisclass * period, 
            data = cvb4064, family = poisson)
dr <- drop1(fitt, test = "Chisq")
##knitr::kable(dr)
xtable(dr,label = "tab:41cumswtab", caption = "Anova, cardiovascular deaths, 40-64.")
@

See Table~\ref{tab:41cumswtab}.

@@

<<41cumswold, fig.cap = cap22, fig.scap = scap22, fig.height = 8>>=
##source("R/plotED.R")
oldpar <- par(mfrow = c(2, 1))
woutp <- plotED.glm(resw, col = scol, logScale = FALSE, relative = TRUE,
                    ylab = "H(90|65)",  main = "Women 65-89", exclude = 3:4)
moutp <- plotED.glm(resm, col = scol, logScale = FALSE, relative = TRUE,
                    ylab = "H(90|65)", main = "Men 65-89")
par(oldpar)
@

<<41cumswoldtab,results='asis'>>=
fitt <- glm(event ~ offset(log(exposure)) + urban + civst + age + sex * period *hisclass,
                ##sex * period + hisclass * period, 
            data = cvb6589, family = poisson)
dr <- drop1(fitt, test = "Chisq")
##knitr::kable(dr)
xtable(dr,label = "tab:41cumswoldtab", caption = "Anova, cardiovascular deaths, 65-90.")
@

See Table~\ref{tab:41cumswoldtab}.

@@

<<12cumsw, fig.cap = cap12, fig.scap = scap12, fig.height = 8>>=
source("R/plotED.glm.R")
oldpar <- par(mfrow = c(2, 1))
woutp <- plotED.glm(resw, col = scol, logScale = FALSE, relative = TRUE, exclude = 3:4,
                    ylab = "TMR(65|40)",  main = "Women 40-64")
moutp <- plotED.glm(resm, col = scol, logScale = FALSE, relative = TRUE, ##exclude = 1:2,
                    ylab = "TMR(65|40)", main = "Men 40-64")
par(oldpar)
@

@@

<<12cumswtab,results='asis'>>=
fitt <- glm(event ~ offset(log(exposure)) + urban + civst + age + sex *hisclass +
                sex * period + hisclass * period, 
            data = cvb4064, family = poisson)
dr <- drop1(fitt, test = "Chisq")
##knitr::kable(dr)
xtable(dr,label = "tab:12cumswtab", caption = "Anova, cancer deaths, 40-64.")
@

See Table~\ref{tab:12cumswtab}.

@@

<<12cumswold, fig.cap = cap22, fig.scap = scap22, fig.height = 8>>=
##source("R/plotED.R")
oldpar <- par(mfrow = c(2, 1))
woutp <- plotED.glm(resw, col = scol, logScale = FALSE, relative = TRUE,
                    ylab = "TMR(90|65)",  main = "Women 65-89", exclude = 3:4)
moutp <- plotED.glm(resm, col = scol, logScale = FALSE, relative = TRUE,
                    ylab = "TMR(90|65)", main = "Men 65-89")
par(oldpar)
@

<<12cumswoldtab,results='asis'>>=
fitt <- glm(event ~ offset(log(exposure)) + urban + civst + age + sex +
                sex * period + hisclass * period, 
            data = cvb6589, family = poisson)
dr <- drop1(fitt, test = "Chisq")
##knitr::kable(dr)
xtable(dr,label = "tab:12cumswoldtab", caption = "Anova, cancer deaths, 65-90.")
@

See Table~\ref{tab:12cumswoldtab}.
% \subsubsection{The effect of income and education on cancer death, 1990--2005}
% 
% Thus, we find indications of social differences in the later decades, particularly between the elite and the workers. The alternative analysis for the last period is based on a separate analysis with
% income and education as explanatory variables.
% 
% <<12lissyoung2>>=
% library(eha)
% library(xtable)
% ##load("data/loTab.rda")
% ##loTab$period <- factor(loTab$period)
% liss6589 <- loTab[loTab$age > 64.5, ]
% @
% 
% 
% The statistical significance of the covariates are shown in Table~\ref{tab:12liswom2} for women and Table~\ref{tab:12lismen2} for men.
% <<12liswom2, results = 'asis'>>=
% fitw2 <- glm(event ~ offset(log(exposure)) + as.factor(age) + urban + period + civst + income + educ, family = poisson, data = liss6589[liss6589$sex == "female", ])
% ##cat("Women:\n")
% drw2 <- drop1(fitw2, test = "Chisq")
% xtable(drw2, caption = "Effect of covariates for women 65-89, 1990-2005, cancer death.", label = "tab:12liswom2")
% ##round(summary(fitw1)$coef[-(1:13), ], 4)
% @
% 
% <<12lismen2, results='asis'>>=
% fitm2 <- glm(event ~ offset(log(exposure)) + as.factor(age) + urban + period + civst + income + educ, family = poisson, data = liss6589[liss6589$sex == "male", ])
% ##cat("\nMen:\n")
% drm2 <- drop1(fitm2, test = "Chisq")
% xtable(drm2, caption = "Effect of covariates for men 65-89, 1990-2005, cancer death.", label = "tab:12lismen2")
% ##round(summary(fitm1)$coef[-(1:13), ], 4)
% dcol <- c("blue", "darkgreen", "darkgreen", "red")
% @
% 
% The estimated effect sizes are shown in Figure \ref{fig:12liswomfig2}
% 
% <<12liswomfig2, fig.cap = "Effect sizes, women 65-89, 1990-2005, cancer death.", fig.height = 4>>=
% #source("R/plotWEduc.R")
% plotWEduc(fitw2)
% @
% 
% 
% And for men:
% 
% %% <<12lismen2, results='asis'>>=
% %% fitm2 <- glm(event ~ offset(log(exposure)) + as.factor(age) + urban + period + civst + income + educ, family = ##poisson, data = liss6589[liss6589$sex == "male", ])
% %% ##cat("\nMen:\n")
% %% drm2 <- drop1(fitm2, test = "Chisq")
% %% xtable(drm2, caption = "Effect of covariates for men 65-89, 1990-2005, cancer death.")
% %% round(summary(fitm1)$coef[-(1:13), ], 4)
% %% dcol <- c("blue", "darkgreen", "darkgreen", "red")
% %% @
% 
% <<12lismanfig2, fig.cap = "Effect sizes, men 65-89, 1990-2005, cancer death.", fig.height = 4>>=
% #source("R/plotWEduc.R")
% plotWEduc(fitm2)
% @
% 
% The social inequality in cancer mortality is confirmed by the approach of analysing the effect of education and income. We find the same consistent pattern as we have found for all-cause mortality and cardiovascular mortality for this period. Higher levels of education and more income is associated with lower cancer mortality. We can therefore conclude that both education and income are strongly decisive aspects of social differentiation when it comes to mortality during the last decades.

@@

<<tableeduc,results='asis'>>=
fit <- coxreg(Surv(enter, exit, event) ~ educ + civst + urban + sex, data = educ[!is.na(educ$civst), ], 
              hazards = FALSE)
dr <- drop1(fit, test = "Chisq")
ltx(fit, dr = dr, caption = "Effect of education on mortality in ages 40-90, 1990-2013.", 
    label = "tab:tableeduc", digits = 4)
@

See Table~\ref{tab:tableeduc}.

@@

<<tableeducheart,results='asis'>>=
fit <- coxreg(Surv(enter, exit, heart) ~ educ + civst + urban + sex, data = educ[!is.na(educ$civst), ], 
              hazards = FALSE)
dr <- drop1(fit, test = "Chisq")
ltx(fit, dr = dr, caption = "Effect of education on cardiovascular mortality in ages 40-90, 1990-2013.", 
    label = "tab:tableeducheart", digits = 4)
@

See Table~\ref{tab:tableeduc}.

@@

<<tableeduccancer,results='asis'>>=
fit <- coxreg(Surv(enter, exit, cancer) ~ educ + civst + urban + sex, data = educ[!is.na(educ$civst), ], 
              hazards = FALSE)
dr <- drop1(fit, test = "Chisq")
ltx(fit, dr = dr, caption = "Effect of education on cancer mortality in ages 40-89, 1990-2013.", 
    label = "tab:tableeduccancer", digits = 4)
@

See Table~\ref{tab:tableeduc}.

@@

<<tableinc>>=
fit <- coxreg(Surv(enter, exit, event) ~ income + civst + urban + sex, data = income[!is.na(income$civst), ],
              hazards = FALSE)
dr <- drop1(fit, test = "Chisq")
@

<<tableincpr,results='asis'>>=
ltx(fit, dr = dr, caption = "Effect of income on mortality in ages 40-89, 1986-2005.", 
    label = "tab:tableincpr", digits = 4)
@

See Table~\ref{tab:tableincpr}.

@@

%The overall pattern is the same for cardiovascular mortality, high income pays off.
<<plotincheart, fig.cap="Cumulative hazards by income and sex, cardiovascular death, 1986-2005.">>=
oldpar <- par(mfrow = c(1, 2))

@@

<<tableincheart>>=
fit <- coxreg(Surv(enter, exit, heart) ~ income + civst + urban + sex, data = income[!is.na(income$civst), ],
              hazards = FALSE)
dr <- drop1(fit, test = "Chisq")
@

<<tableincprheart,results='asis'>>=
ltx(fit, dr = dr, caption = "Effect of income on cardiovascular mortality in ages 40-89, 1986-2005.", 
    label = "tab:tableincpr", digits = 4)
@

See Table~\ref{tab:tableincpr}.

@@

<<tableinccancer>>=
fit <- coxreg(Surv(enter, exit, heart) ~ income + civst + urban + sex, data = income[!is.na(income$civst), ],
              hazards = FALSE)
dr <- drop1(fit, test = "Chisq")
@

<<tableincprcancer,results='asis'>>=
ltx(fit, dr = dr, caption = "Effect of income on cancer mortality in ages 40-89, 1986-2005.", 
    label = "tab:tableincprcancer", digits = 4)
@

See Table~\ref{tab:tableincpr}.
